Synthesising Recursive Functions for\\ First-Order Model Counting

### 2023-03-14 doc/talks/3_long/talk.tex
Describe one more instance
Additions and improvements to the slides

\documentclass{beamer}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{complexity}
\usepackage[style=authoryear]{biblatex}
\usepackage{forest}
\usepackage{soul}
\usepackage{csquotes}
\usepackage{pifont}
\usepackage{venndiagram}

@@

@@

\definecolor{color1}{HTML}{1B9E77}
\definecolor{color2}{HTML}{7570b3}
\definecolor{color3}{HTML}{e7298a} % TODO: I don't like pink
\definecolor{forestgreen}{HTML}{009B55}

@@

% 15 min (8 slides of content) introduction
% 25 min presentation
% 5 min summary
% 20-25 slides
% before it was 15 slides and 20 min
% make sure there's a clear and coherent storyline

\author{\textbf{Paulius Dilkas} \inst{1} \and Vaishak Belle \inst{2}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{\inst{1} National University of Singapore, Singapore \and \inst{2} University of Edinburgh, UK}
\date{TODO: date}

@@

\begin{frame}{What Computers Can and Cannot Do}
  \begin{tikzpicture}[remember picture,overlay]
    \node[right=0.7cm] at (current page.west) (stickman) {\includegraphics[height=0.5\textheight]{stickman.png}};
    \node<-4>[left,shift=({0, -2})] at (current page.east) (computer) {\includegraphics[height=0.5\textheight]{happy_computer}};
    \node<2>[ellipse callout, draw, right=0cm, callout relative pointer={(-1, -0.5)}] at (stickman.north east) {$\int_{-\infty}^{\infty} e^{-t^{2}}\,\mathrm{d}t$};
    \node<2>[ellipse callout, draw, left=-0.5cm, below=0.5cm] at (computer.north west) {$\sqrt{\pi}$};
    \node<3>[ellipse callout, draw, right=-1cm, callout relative pointer={(-2.5, -0.5)}] at (stickman.north east) {Produce a schedule for the nurses at NUH.};
    \node<3>[ellipse callout, draw, below=0.5cm, shift=({-0.5,0}), callout relative pointer={(0.5, -0.25)}] at (computer.north west) {\includegraphics[width=0.1\textwidth]{schedule}};
    \node<4>[ellipse callout, draw, right=0cm, callout relative pointer={(-1, -0.15)}] at (stickman.north east) {Describe this picture. \includegraphics[width=0.25\textwidth]{cat.jpg}};
    \node<4>[ellipse callout, draw, below=0.5cm, shift=({-0.5,0.5}), callout relative pointer={(0.5, -0.5)}] at (computer.north west) {It's a cat with its eyes closed.};
    \node<5>[ellipse callout, draw, right=0cm, callout relative pointer={(-1, -0.15)}, text width=5cm] at (stickman.north east) {If I shuffle a deck of \alert{$n$} cards, how many possible outcomes are there?};
    \node<5>[left,shift=({-1, -1})] at (current page.east) (computer2) {\includegraphics[height=0.5\textheight]{overheating_computer}};
    \node[shift=({1.3, 0.1})] at (current page.south west) {\tiny \color{lightgray}{Vector art by Vecteezy.com}};
  \end{tikzpicture}
\end{frame}

@@

% TODO: (maybe) add a slide about other approaches to WFOMC, making it crystal clear what my contributions are

\begin{frame}{(Unweighted) First-Order Model Counting}
  \begin{columns}[T]
    \begin{column}{0.5\textwidth}
      \begin{itemize}
        \item Example formula: $\forall x \in \Delta\text{. } \pP(x) \lor \qQ(x)$
        \item Let $\Delta \coloneqq \{\, 1, 2 \,\}$
        \item Interpretations: all subsets of $\{\, \pP(1), \qQ(1), \pP(2), \qQ(2) \,\}$
        \item Models:
        \[
          \begin{matrix}
            \{\, \pone, \ptwo \,\}, & \{\, \pone, \qtwo \,\}, & \{\, \pone, \pqtwo \,\},\\
            \{\, \qone, \ptwo \,\}, & \{\, \qone, \qtwo \,\}, & \{\, \qone, \pqtwo \,\},\\
            \{\, \pqone, \ptwo \,\}, & \{\, \pqone, \qtwo \,\}, & \{\, \pqone, \pqtwo \,\}\\
          \end{matrix}
        \]
        \item Each 1-ary predicate...
      \end{itemize}
    \end{column}
    \begin{column}{0.5\textwidth}
      \centering
      \begin{overlayarea}{\textwidth}{4cm}
        \only<1>{
          \begin{venndiagram2sets}[labelNotAB={$\Delta$},labelA={$\pP$},labelB={$\qQ$}]
          \end{venndiagram2sets}
        }%
        \only<2>{
          \begin{venndiagram2sets}[labelNotAB={$\Delta$},labelA={$\pP$},labelB={$\qQ$}]
            \fillOnlyA;
          \end{venndiagram2sets}
        }%
        \only<3>{
          \begin{venndiagram2sets}[labelNotAB={$\Delta$},labelA={$\pP$},labelB={$\qQ$}]
            \fillOnlyB;
          \end{venndiagram2sets}
        }%
        \only<4>{
          \begin{venndiagram2sets}[labelNotAB={$\Delta$},labelA={$\pP$},labelB={$\qQ$}]
            \fillACapB;
          \end{venndiagram2sets}
        }%
        \only<5>{
          \begin{venndiagram2sets}[labelNotAB={$\Delta$},labelA={$\pP$},labelB={$\qQ$}]
            \fillA;
            \fillB;
          \end{venndiagram2sets}
        }%
      \end{overlayarea}
    \end{column}
  \end{columns}
\end{frame}
% TODO: mention that n-ary predicates ~= relations

@@

\begin{frame}{An Example Counting Problem}
  Suppose this room has \alert{$n$} seats, and there are \alert{$m \le n$}
  people in the audience. How many ways are there to seat everyone?
  \vfill

@@

\begin{frame}{Back to Our Example}
  For instance, the following function counts injections
  \[
  f(m, n) =
  \begin{cases}
    1 & \text{if } m = 0 \text{ and } n = 0 \\
    0 & \text{if } m > 0 \text{ and } n = 0 \\
    f(m, n-1) + mf(m-1, n-1) & \text{otherwise.}
  \end{cases}
  \]
  \pause
  \begin{itemize}
    \item \alert{$f$} can be computed in \alert{$\Theta(mn)$} time (via dynamic
          programming).
    \item Optimal time complexity to compute \alert{$n^{\underline{m}}$} is
          \alert{$\Theta(m)$}.
    \item But \alert{$\Theta(mn)$} is still much better than translating to
          propositional logic and running a WMC algorithm.
    \item The rest of this talk is about how such functions can be found
          automatically.
  \end{itemize}
\end{frame}

\begin{frame}{First-Order Knowledge Compilation with \textsc{ForcLift}}
  \begin{columns}
    \begin{column}{0.25\textwidth}
      \centering
      \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-latex},sibling distance=10mm,level distance=10mm]
        \node[fill=red!50,fill on=<2>] (dr) {}
        child {node {}
          child {node[fill=red!50,fill on=<2>] {}
            child {node {}
              child {node {}}
              child {node[fill=red!50,fill on=<2>] (ref) {}}
            }}};
        \draw[-latex, bend right,alt=<2>{red}{black}] (ref) to (dr);
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.75\textwidth}
      \begin{block}{Workflow Before}
        \begin{enumerate}
        \item Compile the formula to a \alert{circuit}
        \item Evaluate to get the answer
        \end{enumerate}
      \end{block}
      \pause
      \begin{block}{Workflow After}
        \begin{enumerate}
        \item Compile the formula to a \alert{graph}
        \item Extract the definitions of functions
        \item Simplify
        \item Supplement with \alert{base cases}
        \item Evaluate to get the answer
        \end{enumerate}
      \end{block}
    \end{column}
  \end{columns}
\end{frame}
% Only Step 1 is fully automated at this point. The other steps are executed by hand.
% Two parts: compilation and inference/evaluation.

@@

{ % all template changes are local to this group.
    \setbeamertemplate{navigation symbols}{}
    \begin{frame}<article:0>[plain]
        \begin{tikzpicture}[remember picture,overlay]
            \node[at=(current page.center)] {
                \includegraphics[keepaspectratio, height=\paperheight]{zoom.jpg}
            };
        \end{tikzpicture}
     \end{frame}
}

>>>>

\documentclass{beamer}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{complexity}
\usepackage[style=authoryear]{biblatex}
\usepackage{forest}
\usepackage{soul}
\usepackage{csquotes}
\usepackage{pifont}

@@

\newcommand\blfootnote[1]{%
  \begingroup
  \renewcommand\thefootnote{}\footnote{#1}%
  \addtocounter{footnote}{-1}%
  \endgroup
}

@@

\definecolor{color1}{HTML}{1B9E77}
\definecolor{color2}{HTML}{7570b3}
\definecolor{color3}{HTML}{a6761d}
\definecolor{forestgreen}{HTML}{009B55}

\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}

@@

% 15 min (8 slides, currently have 7) introduction
% 25 min (13 slides, currently have 10) presentation
% 5 min (3 slides, currently have 1) summary
% make sure there's a clear and coherent storyline

\author{Paulius Dilkas \\ \vspace{1cm} \tiny{Joint work with Vaishak Belle (Univeristy of Edinburgh, UK)}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{National University of Singapore, Singapore}
\date{\today}
% TODO: update the date
% TODO: unify the use of colours
% TODO: do I want to mention forclift and/or crane?

@@

\begin{frame}{What Computers Can and Cannot Do}
  \begin{tikzpicture}[remember picture,overlay]
    \node[right=0.7cm] at (current page.west) (stickman) {\includegraphics[height=0.5\textheight]{stickman.png}};
    \node<-4>[left,shift=({0, -2})] at (current page.east) (computer) {\includegraphics[height=0.5\textheight]{happy_computer}};
    \node<2>[ellipse callout, draw, right=0cm, callout relative pointer={(-1, -0.5)}] at (stickman.north east) {$\int_{-\infty}^{\infty} e^{-t^{2}}\,\mathrm{d}t$};
    \node<2>[ellipse callout, draw, left=-0.5cm, below=0.5cm] at (computer.north west) {$\sqrt{\pi}$};
    \node<3>[ellipse callout, draw, right=-1cm, callout relative pointer={(-2.5, -0.5)}] at (stickman.north east) {Produce a schedule for the nurses at NUH.};
    \node<3>[ellipse callout, draw, below=0.5cm, shift=({-0.5,0}), callout relative pointer={(0.5, -0.25)}] at (computer.north west) {\includegraphics[width=0.1\textwidth]{schedule}};
    \node<4>[ellipse callout, draw, right=0cm, callout relative pointer={(-1, -0.15)}] at (stickman.north east) {Describe this picture. \includegraphics[width=0.25\textwidth]{cat.jpg}};
    \node<4>[ellipse callout, draw, below=0.5cm, shift=({-0.5,0.5}), callout relative pointer={(0.5, -0.5)}] at (computer.north west) {It's a cat with its eyes closed.};
    \node<5>[ellipse callout, draw, right=0cm, callout relative pointer={(-1, -0.15)}, text width=5cm] at (stickman.north east) {If I shuffle a deck of \alert{$n$} cards, how many possible outcomes are there?};
    \node<5>[left,shift=({-1, -1})] at (current page.east) (computer2) {\includegraphics[height=0.5\textheight]{overheating_computer}\blfootnote{Terms and conditions apply.}};
    \node[shift=({1.3, 0.1})] at (current page.south west) {\tiny \color{lightgray}{Vector art by Vecteezy.com}};
  \end{tikzpicture}
\end{frame}
% TODO: replace the cat thing with a funny DALL-E picture

@@

% TODO: add a slide about other approaches to WFOMC, making it crystal clear
% what my contributions are. This will finish the introduction

\begin{frame}{(Unweighted) First-Order Model Counting}
  \begin{itemize}
    \item Example formula: \\
          \alert{$\forall x \in \Delta\text{. } \pP(x) \lor \qQ(x)$}
    \item Let \alert{$\Delta \coloneqq \{\, 1, 2 \,\}$}
    \item Interpretations: all subsets of\\
          \alert{$\{\, \pP(1), \qQ(1), \pP(2), \qQ(2) \,\}$}
    \item<2-> Models:
    \[
      \begin{matrix}
        \{\, \pone, \ptwo \,\}, & \{\, \pone, \qtwo \,\}, & \{\, \pone, \pqtwo \,\},\\
        \{\, \qone, \ptwo \,\}, & \{\, \qone, \qtwo \,\}, & \{\, \qone, \pqtwo \,\},\\
        \{\, \pqone, \ptwo \,\}, & \{\, \pqone, \qtwo \,\}, & \{\, \pqone, \pqtwo \,\}\\
      \end{matrix}
    \]
  \end{itemize}
  \onslide<3->{
    \begin{block}{Intuition}
      \begin{itemize}
        \item Each 1-ary predicate is like a \alert{subset}
        \item For \alert{$n>1$}, each \alert{$n$}-ary predicate is like a
              \alert{relation}
        \item FOMC counts \alert{combinations of relations}
      \end{itemize}
    \end{block}
  }
  \begin{tikzpicture}[remember picture,overlay,visible on=<2->]
    \tikzset{shift={(current page.north east)}}
    \tikzset{shift={(-4, -2.9)}}
    \node[draw,circle,minimum size=3cm,label={135:$\pP$},fill=color1] (P) at (0,0) {};
    \node[draw,circle,minimum size=3cm,label={45:$\qQ$},fill=color2] (Q) at (1.8,0) {};
    \begin{scope}
      \clip (0,0) circle(1.5cm);
      \clip (1.8,0) circle(1.5cm);
      \fill[color3](0,0) circle(1.5cm);
    \end{scope}
    \draw (0,0) circle(1.5cm);
    \draw (1.8,0) circle(1.5cm);
  \end{tikzpicture}
\end{frame}

@@

\begin{frame}{A (Slightly) More Complicated Example}
  Suppose this room has \alert{$n$} seats, and there are \alert{$m \le n$}
  people in the audience. How many ways are there to seat everyone?
  \vfill

@@

% TODO: it would be good to have a similar compilation walkthrough for an instance with Ref as well

\begin{frame}{Back to Our Example}
  The following function counts injections:
  \[
  f(m, n) =
  \begin{cases}
    1 & \text{if } m = 0 \text{ and } n = 0 \\
    0 & \text{if } m > 0 \text{ and } n = 0 \\
    f(m, n-1) + mf(m-1, n-1) & \text{otherwise.}
  \end{cases}
  \]
  \pause
  \begin{itemize}
    \item \alert{$f(m, n)$} can be computed in \alert{$\Theta(mn)$} time
    \begin{itemize}
      \item using dynamic programming
    \end{itemize}\pause
    \item \alert{Optimal} time complexity to compute \alert{$n^{\underline{m}}$}
          is \alert{$\Theta(m)$}
    \item But \alert{$\Theta(mn)$} is still much better than translating to
          propositional logic and running a WMC algorithm \pause
    \item The rest of this talk is about how to find such functions automatically
  \end{itemize}
\end{frame}

\begin{frame}{First-Order Knowledge Compilation: Before and After}
  \centering
  \begin{tikzpicture}
    \node (start) {\tiny $\begin{gathered}
        \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }\pP(x, y)\\
        \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }\pP(x, y) \land \pP(x, z) \Rightarrow y=z\\
        \forall w, x \in \Gamma\text{. }\forall y \in \Delta\text{. }\pP(w, y) \land \pP(x, y) \Rightarrow w=x
      \end{gathered}$};
    \node[draw,rounded rectangle,fill=color2!50,below=0.3cm of start] (kc) {Compilation};
    \node[below=0.09cm of kc] (graph) {
      \scalebox{.5}{
      \begin{forest}
        for tree={sn edges,grow=0,reversed,fill=red!50}
        [,ellipse,draw,name=gdr
        [,ellipse,draw
        [,ellipse,draw
        [,ellipse,draw
        [,ellipse,draw]
        [,ellipse,draw,name=ref]
        ]
        ]
        ]
        ]
        \draw[-Latex,bend left=20] (ref) to (gdr);
      \end{forest}
      }
    };
    \node[draw,rounded rectangle,fill=color2!50,below=0.1cm of graph] (interpretation) {Conversion};
    \node[below=0.3cm of interpretation] (initial) {\tiny $f(m, n) = \sum_{l=0}^m \binom{m}{l} [l<2] \times f(m-l, n-1)$};
    \node[draw,rounded rectangle,fill=color2!50,below=0.3cm of initial] (simplification) {Simplification};
    \node[below=0.3cm of simplification] (simplified) {\tiny $f(m, n) = f(m, n-1) + mf(m-1, n-1)$};
    \node[draw,rounded rectangle,fill=color2!50,below=0.3cm of simplified] (evaluation) {Evaluation};
    \node[below=0.3cm of evaluation] (end) {\includegraphics[height=0.4cm]{42.png}};
    \node[right=0.3cm of evaluation] (sizes) {\tiny $\begin{aligned}
        f(0,0) = 1&\text{{,} }f(m, 0) = 0\\
        m = 7&\text{{,} }n=2
      \end{aligned}$};
    %\node at (0.5, -0.55) {\textsc{Crane}};
    %\node[draw,fit={(kc) (interpretation)},inner ysep=6pt,yshift=5pt] {};
    \draw[-Latex,color=color2] (start) -- (kc);
    \draw[-Latex,color=color2] (kc) -- ($(graph)+(0,0.2)$);
    \draw[-Latex,color=color2] ($(graph)+(0,-0.3)$) -- (interpretation);
    \draw[-Latex,color=color2] (interpretation) -- (initial);
    \draw[-Latex,color=color2] (initial) -- (simplification);
    \draw[-Latex,color=color2] (simplification) -- (simplified);
    \draw[-Latex,color=color2] (simplified) -- (evaluation);
    \draw[-Latex,color=color2] (evaluation) -- (end);
    \draw[-Latex,color=color2] (sizes) -- (evaluation);
  \end{tikzpicture}
\end{frame}
% TODO: optimize edge lengths
% TODO: add forclift on the left, with its own example instance
% TODO: check if the math works out

% TODO: talk #4 might have some useful additions to this section

@@

% TODO: maybe replace the zoom picture with some galaxy picture?
{ % all template changes are local to this group.
    \setbeamertemplate{navigation symbols}{}
    \begin{frame}<article:0>[plain]
        \begin{tikzpicture}[remember picture,overlay]
            \node[at=(current page.center)] {
                \includegraphics[keepaspectratio, height=\paperheight]{zoom.jpg}
            };
        \end{tikzpicture}
     \end{frame}
}
